---
title: "Appendix S1"
author: "Buckley et al."
output:
  html_document: default
  pdf_document: default
  word_document: default
  rmarkdown::html_vignette: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
##Load TrenchR package
library(TrenchR)
library(maptools)
library(ggplot2)
library(reshape)
```

# APPENDIX S1
We provide an additional example using microclimate data measured along an elevation gradient in CO, USA to illustrate use of TrenchR. We first use a time series of air temperatures and wind speeds collected at multiple heights to examine profiles and surface roughness. We then use a time series of air and surface temperatures, wind speeds, and solar radiation collected at a single height to implement a biophysical model for grasshoppers and compare estimates to observations.

# Microclimate models
```{r, include=FALSE}
setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/data/")
C1all=read.csv("microclimate_C1.csv", sep=",", na.strings = c("-49.99","NA"))
```

```{r, include=FALSE}
#ANALYZE SURFACE ROUGHNESS AND PROFILES
C1all$day=as.POSIXlt(as.character(C1all$datetime), tz='MST', format="%m/%d/%Y %H:%M")$mday
C1all$hour=as.POSIXlt(as.character(C1all$datetime), tz='MST', format="%m/%d/%Y %H:%M")$hour 

C1.dayhr<- aggregate(C1all[,c(2:8,13:22)], by=list(C1all$day, C1all$hour), FUN=mean, na.rm=TRUE)
names(C1.dayhr)[1:2]=c("day","hour")
C1.hr<- aggregate(C1all[,c(2:8,13:22)], by=list(C1all$hour), FUN=mean, na.rm=TRUE)
names(C1.hr)[1]="hour"

C1.hr= as.matrix(C1.hr)
```

We first examine how temperature and wind speed profiles can be used to scale conditions to the heights of organisms. We first use wind speed variation across hours and heights to estimate surface roughness.
```{r}
#estimate surface roughness based on wind velocity (m/s) at a vector of reference heights, zr (m)
z0= apply(C1.hr[, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], FUN=surface_roughness, MARGIN=1, zr=c(0.25,0.5,1.0,1.5))

```

We then calculate the temperature and wind speed profiles for a given hour using multiple, alternative functions in TrenchR.
```{r}
#subset to day and hour
C1.sub= subset(C1.dayhr, C1.dayhr$day==10)
hr=10

#vector of heights
z.seq <- seq(0, 2, by = 0.1)

#estimate profiles using three different functions including the assumption of neutral and segmented profiles. 
#T_r	and u_r are temperatures (C) and windspeeds (m/s), respectively, at reference height zr (m).
#z0	is surface roughness (m); z is the height to scale to (m).
#T_s	is surface temperature (C).

#temperature profiles
t.seq1= sapply(z.seq, FUN=air_temp_profile_neutral, T_r=C1.sub[hr,"Air0.5m"], z0=z0[hr], zr=0.5, T_s=C1.sub[hr,"Soil"])

t.seq2= sapply(z.seq, FUN=air_temp_profile, T_r=C1.sub[hr,"Air0.5m"], u_r=C1.hr[hr,"Wind0.5m"], zr=0.5, z0=z0[hr], T_s=C1.sub[hr,"Soil"])

t.seq3= sapply(z.seq, FUN=air_temp_profile_segment, T_r=C1.sub[hr, c("Air0.25m","Air0.5m","Air1.0m","Air1.5m")],u_r=C1.sub[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.25,0.5,1,1.5), z0=rep(z0[hr],4), T_s=C1.sub[hr,"Soil"])

#windspeed profiles
u.seq1= sapply(z.seq, FUN=wind_speed_profile_neutral, u_r=C1.sub[hr,"Wind0.5m"], zr=0.5, z0=z0[hr])

#u.seq2= sapply(z.seq, FUN=wind_speed_profile, u_r=C1.sub[hr,"Wind0.5m"], zr=0.5, z0=z0[hr])

u.seq3= sapply(z.seq, FUN=wind_speed_profile_segment, u_r=C1.sub[hr, c("Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.5,1,1.5), z0=rep(z0[hr],3))
```

The temperature and wind speed profiles illustrate how temperatures generally decrease and wind speeds generally increase with height above the ground (Figure 1). Profiles can be fit as a single smooth function or segments.
```{r, echo=FALSE}
#replace infinity with N
t.seq1[which(!is.finite(t.seq1))]=NA

plot(t.seq1, z.seq, type = "l", xlab = "Temperature (C)", ylab = "z (height above ground in m)")
points(t.seq2, z.seq, type = "l", lty="dashed")
points(t.seq3, z.seq, type = "l", lty="dotted")
points(C1.sub[hr, c("Air0.25m","Air0.5m","Air1.0m","Air1.5m")], c(0.25,0.5,1,1.5), type = "p", col="red")

plot(u.seq1, z.seq, type = "l", xlab = "windspeed (m/s)", ylab = "z (height above ground in m)")
#points(u.seq2, z.seq, type = "l", lty="dashed")
points(u.seq3, z.seq, type = "l", lty="dotted")
points(C1.sub[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], c(0.25,0.5,1,1.5), type = "p", col="red")
```

*Figure 1. Temperature (top) and wind speed (bottom) profiles using similar continuous [solid: air_temp_profile_neutral(); dashed: air_temp_profile()] and segmented estimation [dotted: air_temp_profile_segment()].*

# Biophysical models
```{r, include=FALSE}
#Grasshopper data
setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/data/")
spec.dat=read.csv("SpecData.csv")

#read weather data
dat=read.csv("DatAll_Pace2011.csv", sep=",", na.strings = c("-49.99","NA"))
#match site data
sites=read.csv("SiteElevations.csv", sep=",")

#combine soil temperatures
dat$SoilTemp= rowMeans(dat[,c("SoilTemp","SoilTemperature")], na.rm = TRUE) 

#omit problematic temperatures
dat$Temp[dat$Temp< -10]=NA

#add elevation
match1= match(dat$site, sites$Site)
dat$Elev=sites[match1,"Elevation"]

#calculate Julian day
date.ct <- strptime(dat$datetime, format="%m/%d/%Y %H:%M", tz="MST")
dat$J= round(as.numeric(julian(date.ct, origin = as.Date("2011-01-01") )))
dat$hour= date.ct$hour

#match site data
match1= match(dat$site, sites$Site) 
dat$lon= sites$Lon[match1]
dat$lat= sites$Lat[match1]

#Subset to August
dat.dt<- strptime(dat$datetime, format="%m/%d/%Y %H:%M", tz="MST") #as.Date
dataug= subset(dat, dat.dt$mon==7)
#subset to day
dat= subset(dataug, dataug$Rad>5)
```

We next illustrate quantifying operative environmental temperatures, which are estimated body temperatures for grasshoppers in the specified environments. We first read and process time series of environmental data. We estimate the zenith angle, the angle between the sun's rays and the vertical, across Julian days and hours.
```{r}
#Calculate zenith
dat$psi= zenith_angle(dat$J,lat=dat$lat, lon=dat$lon, hour=dat$hour)

```

We scale the air temperature and wind speed from a reference height of 0.5m to grasshopper height (0.001m).
```{r}
z<-0.001 #specify distance from ground 

#scale temperature T_r from reference height zr=0.5m to grasshopper height based on windspeed u_r, surface temperature T_s, and surface roughness z0.
dat$Tgrass <- air_temp_profile(T_r=dat$Temp, u_r=dat$Wind, zr=0.5, z0=0.02, z=z, T_s=dat$SoilTemp)

#scale wind to grasshopper height
dat$Ugrass= wind_speed_profile_neutral(u_r=dat$Wind, zr=0.5, z0=0.02,z=z)
```

```{r, include=FALSE}
#cut nas
dat= dat[,c("Rad","Temp","Wind","SoilTemp","SanguinipesTemp","Elev","J", "hour","psi","Tgrass","Ugrass")]
dat= na.omit(dat)
```

We then use biophysical models to estimate body temperatures based on air and surface temperatures, wind speed, and solar radiation. We first use a biophysical model specifically configured for grasshoppers.
```{r, warning=FALSE}

##Calculate operative temperature, Te
dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Ugrass, H=dat$Rad, K_t=1, psi=dat$psi, l=0.0211, Acondfact=0.0, abs=0.9, r_g=0.5) 

#T_a	and T_g are air and surface temperatures (C), respectively; u	is wind speed (m/s); 
#H is total (direct + diffuse) solar radiation flux (W/m^2); 
#K_t	is the clearness index (dimensionless), which is the ratio of the global solar radiation measured at the surface to the total solar radiation at the top of the atmosphere;
#psi is solar zenith angle (degrees); l	is grasshopper length (m);
#Acondfact is the proportion of the grasshopper surface area in contact with the ground;
#abs is absorptivity of grasshopper to solar radiation (proportion);
#r_g is substrate solar reflectivity (proportion).
```

```{r, include=FALSE}
#fix infinite
dat$Te[is.infinite(dat$Te)]<-NA
#ditch big numbers for now
dat$Te[dat$Te>100]<-NA
```

We then implement a generic energy budget. We first illustrate the estimation of convective heat transfer coefficients as a function of wind speed.
```{r}
#Estimate heat transfer coefficient using empirical relationship
dat$H_L=heat_transfer_coefficient(V=dat$Ugrass,D=0.0211,K=0.025, nu= 15.3 * 10^(-6) , taxon="cylinder")
#where V is	air velocity (m/s); D is the characteristic dimension (m);
#K is the thermal conductivity of air (W m^-1 K^-1);
#nu is the kinematic viscosity of air (m^2 s^-1)

#Also estimate using a spherical approximation and a simplified version of the approximation.
H_L1= heat_transfer_coefficient_approximation(V=dat$Ugrass,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxon="flyinginsect")
H_L2= heat_transfer_coefficient_simple(V=dat$Ugrass,D=0.0211, type="Gates")

#surface area from length(m)
A= surface_area_from_length(l=0.0211)
```

We then implement the energy budget as follows.
```{r, results = FALSE, warning=FALSE}
TeGates= rep(NA, nrow(dat) )

for(ind in 1:nrow(dat) ){

  out= try(Tb_Gates(A=A, D=0.0211/3, psa_dir=0.5, psa_ref=1-0.5, psa_air=1, psa_g=0.0, T_g=dat$SoilTemp[ind]+273, T_a=dat$Tgrass[ind]+273, Qabs=A*0.66*dat$Rad[ind], epsilon=0.95, H_L=dat$H_L[ind], ef=1.3, K=0.15), silent=TRUE)
  if(is.numeric(out))TeGates[ind]=out
}
#where psa_dir, psa_ref, psa_air, psa_g are the proportions surface area exposed to direct radiation from the sky (or enclosure), reflected radiation from the ground, and air, and in contact with the ground, respectively;
#Qabs is the solar and thermal radiation absorbed (W);
#epsilon is longwave infrared emissivity of skin (proportion);
#H_L is the convective heat transfer coefficient (W m^-2 K^-1);
#ef	is the enhancement factor, used to adjust H_L to field conditions;
#K is thermal conductivity. 
```

We aggregate the data and plot average conditions across sites (Figure 2). The simplifying assumptions in the biophysical model implementation reduce alignment with the observed temperatures of grasshoppers, but the divergence of grasshopper temperatures from air temperatures at 2m and, to a lesser extent, air temperatures at grasshopper height demonstrate the importance of implementing biophysical models.
```{r, echo=FALSE, warning=FALSE}
dat$TeGates=TeGates-273.15

#drop NAs
dat= dat[which(!is.na(dat$TeGates)),]

#aggregate by hour
dat.h= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour), FUN=mean, na.rm=TRUE))
#aggregate by day and hour
dat.dh= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour, dat$J), FUN=mean, na.rm=TRUE))

#convert to long format
dat.s= melt(dat.h, id = c("Elev", "J","hour","psi"))

#plot
dat.pl= subset(dat.s, dat.s$variable %in% c("Temp","SoilTemp","SanguinipesTemp","Te","Tgrass","TeGates") )
dat.pl$variable= as.character(dat.pl$variable)

#change names
dat.pl[which(dat.pl$variable=="Temp"),"variable"]<-"Ta"
dat.pl[which(dat.pl$variable=="SoilTemp"),"variable"]<-"Ts"
dat.pl[which(dat.pl$variable=="SanguinipesTemp"),"variable"]<-"Te observed"
dat.pl[which(dat.pl$variable=="Te"),"variable"]<-"Te"
dat.pl[which(dat.pl$variable=="Tgrass"),"variable"]<-"Ta grasshopper height"
dat.pl[which(dat.pl$variable=="TeGates"),"variable"]<-"Te Gates"

#make factor
dat.pl$variable= factor(dat.pl$variable, levels=c("Ta","Ts","Ta grasshopper height","Te","Te Gates","Te observed"))
#make label
dat.pl$elev.lab= paste(dat.pl$Elev,"m", sep=" ")
#make Ta, Te classification
dat.pl$Ttype="input"
dat.pl$Ttype[which(dat.pl$variable %in% c("Te","Te Gates","Te observed") )]="estimates"

ggplot(data=dat.pl, aes(x=hour, y = value, color=variable))+geom_line(aes(lty=variable) )+facet_wrap(~elev.lab)+theme_classic()+ theme(legend.position="bottom")+ylim(15,65)

```

*Figure 2. Diurnal temperature variation averaged across days in August varies across the elevations of CO sites (panels) and the type of temperature (C). Microclimate models indicate that the air temperatures experienced by grasshoppers near the ground (Ta grasshopper height) is generally above air temperatures (Ta) and close to surface temperatures (Ts). Body temperatures estimated by a grasshopper specific biophysical model (Te) are fairly similar to but sometimes less than the body temperatures of a physical grasshopper model implanted with a thermistor (Te observed). The general biophysical models that makes simplifying assumptions (Te Gates) tends to overestimate body temperatures.*

```{r, include=FALSE}
#plot by date time
dat.dh$Jhr= dat.dh$J +dat.dh$hour/24

#convert to long format
dat.s= melt(dat.dh, id = c("Elev", "J","hour","psi","Jhr"))

#plot
dat.s= subset(dat.s, dat.s$variable %in% c("Temp","SoilTemp","SanguinipesTemp","Te","Tgrass") )
#,"TeGates"

plot.t= ggplot(data=dat.s, aes(x=Jhr, y = value, color=variable))+geom_line(aes(lty=variable))+facet_grid(Elev~1)
```

```{r, include=FALSE}
setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/figures/")

pdf("TeTime2.pdf", height = 10, width = 15)
plot.t
dev.off()

```
