---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abby Meyer, Yutaro Sakairi,..."
output:
  html_document: default
  word_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: methods-in-ecology-and-evolution.csl
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

-   Within the word count (6000-7000 words for Standard Articles, 3000
    words for Applications and Practical Tools)

-   Continuous line and page numbering throughout

Lauren B. Buckley^1^, Bryan Briones Ortiz^1^, Isaac Caruso^1^, Aji
John^1^, Ofir Levy^2^, Abby Meyer^1^, Yutaro Sakairi^1^,...

1.  Department of Biology, University of Washington, Seattle, WA
    98195-1800, USA
2.  

Corresponding Author: Lauren Buckley,
[lbuckley\@uw.edu](mailto:lbuckley@uw.edu){.email}

Running head: TrenchR microclimate and biophysical models 

Type: applications

## Abstract

1.  Much understanding of climate change responses relies on the
    assumption that organismal body temperatures are equal to temporally
    averaged air temperatures high above the ground. However, most
    organisms live near the ground and acute exposure to solar radiation
    can substantially elevate their body temperatures.
2.  We introduce the TrenchR package, which aids in Translating
    Environmental Change into organismal responses
    (<https://github.com/trenchproject/TrenchR>). The package includes
    microclimate models to vertically scale weather station data to
    organismal heights. The biophysical modelling tools include both
    general models for heat flows and specific models to predict body
    temperatures for a variety of ectothermic taxa. Additional functions
    model and temporally partition air and soil temperatures and solar
    radiation.
3.  [Main results]
4.  The package aims to introduce and enable microclimate and
    biophysical modelling to improve ecological and evolutionary
    forecasting and includes tutorials and well as a series of
    educational modules.

**Keywords**: biophysical, climate change, ecotherm, energy budget,
mechanistic models, microclimate, scaling, temperature

## Introduction

Many responses of organisms and ecosystems to climate change are
inconsistent with current predictive models [@maguire2015modeling]. Some
of these predictive shortcomings stem from omitting spatial and temporal
environmental variation and how it interacts with organismal phenotypes
(e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
Many analyses assume that organismal body temperatures are equal to air
temperate at weather station height (usually 2m). However, air
temperatures near the ground where most organisms reside can be
considerably warmer and absorption of solar radiation can raise body
temperatures well above air temperatures, resulting in potential
discrepancies of tens of degrees C [@sinclair2016; @sunday2014;
@kearney2009]. Acute thermal stress events are often more relevant to
the physiology and energy balance of organisms than daily, monthly, or
annually aggregated environmental conditions. Yet, daily and seasonal
environmental variation is often neglected [@dillon2016introduction;
@dillon2016life]. We aim to help remedy these shortcomings by
introducing the TrenchR package as a component of the TrEnCh project
(<https://trenchproject.github.io>), which builds computational and
visualization tools to Translate Environmental Change into organismal
responses.

The mathematical tools for leveraging environmental data to predict
organismal conditions have long lingered in books
[@campbell2000introduction; @gates1980biophysical] and articles
[@porter1969thermodynamic; @porter1983biophysical], but adoption of the
tools has not kept pace with research on climate change responses.
Adoptions of biophysical approaches are expanding with recognition of
the importance of spatial and temporal environmental variation to
determining climate change responses [@buckley2018leveraging;
@kearney2017nichemapr]. Increasing dissemination of computational tools
is facilitating adoption. The TrenchR package is intended to complement
the NicheMapR package, which includes sophisticated microclimate
[@kearney2017nichemapr] and biophysical [@kearney2020] models. Many of
the approaches are similar between the packages, but TrenchR focuses on
simple functions to aid the understanding and transparency of
biophysical approaches. The TrenchR functions allow for modular use and
easy adaptation to a variety of organismal and research questions. Some
energy budget calculation components of biophysical models are also
available in the ThermImage package [@tattersall2017]. Although many of
the TrenchR functions are general, some components of TrenchR are best
suited for ectothermic animals. We have minimized inclusion of plant
specific biophysical models since they are the focus of several R
packages [tealeaves: [@muir2019tea]; plantecophys:
[@duursma2015plantecophys]].

## Methods and features

The TrenchR package aims to promote transparency and reproducibility and
we welcome contributions and corrections from users. Our package is
built using the devtools methodology
(<https://github.com/hadley/devtools>) in Github. Issues and feature
requests can be contributed in Github
(<https://github.com/trenchproject/TrenchR/>). The package is maintained
and checked using Travis CI as a tool for continuous integration.

Functions were adapted from biophysical ecology textbooks
[@campbell2000introduction; @gates1980biophysical] and research
articles. The source reference is included in each function's header. We
have incorporated default parameters (e.g., organism emissivity or heat
conduction rates) when general values are available. Many functions
additionally include comments that describe potential parameterizations.
The biophysical ecology texts [@campbell2000introduction;
@gates1980biophysical] include many tables with parameter values and we
provide several tables in TrenchR [Table 8.2XXXX].

### Components

TrenchR functions are organized into the following categories:

**Utility functions** allow estimating environmental metrics that form
the basis of microclimate and biophysical models (e.g., zenith and
declination angles), facilitate modeling diurnal temperature variation,
and estimate biologically relevant aggregate metrics such as degree days
available for development.

**Microclimate functions** allow estimating the environmental conditions
experienced by organisms. Temperature and wind profile functions scale
environmental conditions from sensors to organismal height. Radiation
functions allow estimating incoming solar radiation and partitioning the
variation diurnally and across components (i.e., direct, diffuse,
reflected).

**General biophysical functions** provide generalized models of heat
exchanges between organisms and their environment so that users can
build custom biophysical models. The functions allow implementing an
energy balance of the following components of heat exchange between
organisms and their environment: *radiative heat exchanges* of solar and
thermal radiation; *convective heat exchanges* between organisms and
their surrounding fluid (air or water) driven by fluid flow; *conductive
heat exchanges* between organisms and the ground due to physical
contact; and *evaporative heat exchanges* associated with organisms'
evaporative water loss. Additional functions aggregate these forms of
heat exchange into energy balances and to use the energy balances to
predict body temperatures.

**Specific biophysical functions** provide biophysical models that have
been built for specific organisms based on their physical properties,
behavior, environment, and life history. We currently provide published
biophysical models for lizards, salamanders, butterflies, grasshoppers,
limpets, mussels, and snails. Most models predict operative
environmental temperatures, which are the equilibrium body temperatures
of organisms with specified physical properties in a specific
microenvironment and assume no heat exchange via metabolism or
evaporation [@bakken1992measurement].

### Vignettes and introductory modules

We introduce the function in several vignettes. A good place to start is
the *Allometry and conversions* tutorial, which provides tools for
preparing data such as estimating additional dimensions of organisms
from measured dimensions. The *Estimating microclimates* tutorial
provides resources for estimating the environmental conditions
experienced by organisms. This includes estimating solar radiation and
its components, diurnal variation in temperature and radiation,
temperature and windspeed profiles, and soil temperatures and profiles.
Finally, the core biophysical modelling functions are described in a
tutorial on *Using energy balances to estimate body temperatures*.
Components of an energy budget can be estimated using individual
functions and then body temperatures can be solved for using either a
generic energy balance or taxa specific biophysical models.

For additional background on microclimate and biophysical modelling, we
have updated a series of educational modules entitled "Physical
processes in ecosystems". We intend the modules to provide a more
contained and accessible introduction to microclimate and biophysical
modelling than that included in classic textbooks
[@campbell2000introduction; @gates1980biophysical]. The modules provide
less detailed coverage of many of the topics contained in the books. The
series of 14 modules starts with an overview of the calculus and physics
principles underlying the modelling. Thermodynamics and energy budget
modelling are then introduced with detailed examples. Modules address
the climate space concept, operative temperatures, biophysical models
for leaves and sheep, and heat flow in soils. We have revised the
modules (which originated from an NSF training grant in 1979 and lacked
broad distribution) to include R code and utilization of TrenchR
functions. We expanded the original series of tutorials to include
introductions by Michael Kearney to the Microclim environmental data
[@kearney2014microclim] and integrating the Microclim data with
biophysical modelling to examine broad scale climatic limits. The
modules are provided in html form and available for download as a pdf at
the bookdown server
(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>). The
R markdown files are available in Github at
<https://github.com/trenchproject/TrenchRmodules>.

## Examples

We illustrate the use of the TrenchR package by estimating an energy
budget for a *Sceloporus* lizard on June 1, 2018, in Santa Fe, New
Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also
incorporated in the *Using energy balances to estimate body
temperatures* tutorial. We first estimate the energy budget with
component functions and we then use an integrated biophysical model to
estimate operative environmental temperatures. We start by generating
environmental inputs. See Appendix S1 for another example examining a
time series of microclimate data and applying and testing the
biophysical models.

We will use the energy budget to estimate the lizard's body temperature,
$T_b$, which can depart dramatically from the air temperatures due to
energy exchange with the environment. Heat is gained from absorbing
solar and thermal radiation and from metabolic reactions. Heat is lost
through the organism's emission of radiation and the evaporation of
water. The organism exchanges heat with the surrounding air or water via
convection and with the ground via conduction. The balance of these heat
exchanges can be estimated and often referred to as operative
environmental temperature, $T_e$ [@bakken1992measurement].

Let's assume the lizard is in an unshaded location where a weather
station at standard height (2m) indicates that the daily air temperature
varies from a minimum of 10°C to a maximum of 25°C and the wind speed
averages 1 m/s. The soil surface temperature varies from a minimum of
15°C to a maximum of 30°C. We assume that atmospheric transmissivity
$\tau=0.7$ and albedo $\rho=0.6$

### Environmental data

Let us first prepare the environmental data for analysis. We will need
to estimate hourly air and soil temperatures and radiation. We start by
estimating the day of year and the timing of sunrise and sunset:

```{r, echo=FALSE}
library(TrenchR)

#to be REMOVED
Tb_CampbellNorman=function(T_a, T_g, S, alpha_L=0.96, epsilon=0.96, c_p=29.3, D, V){
    
  stopifnot(T_a>200, T_a<400, epsilon>=0.5, epsilon<=1, c_p>=0, D>0, V>=0)
  
  #Stefan-Boltzmann constant
  sigma= 5.673*10^(-8) #W m^(-2) K^(-4)
  
  #solar and thermal radiation absorbed
  L_a=sigma*T_a^4  # (10.7) long wave flux densities from atmosphere 
  L_g=sigma*T_g^4  # (10.7) long wave flux densities from ground
  F_a=0.5; F_g=0.5 #proportion of organism exposure to air and ground, respectively
  R_abs= S+alpha_L*(F_a*L_a+F_g*L_g) # (11.14) Absorbed radiation
  
  #thermal radiation emitted
  Qemit= epsilon*sigma*T_a^4
  
  #conductance
  g_Ha=1.4*0.135*sqrt(V/D) # boundary conductance, factor of 1.4 to account for increased convection (Mitchell 1976), assumes forced conduction
  g_r= 4*epsilon*sigma*T_a^3/c_p # (12.7) radiative conductance
  
  # operative environmental temperature
  T_e=T_a+(R_abs-Qemit)/(c_p*(g_r+g_Ha))                       

  return(T_e) 
}
```

```{r}
#set up input data as variables
#latitude and longitude (degrees) and elevation (m)
lat=35.69; lon= -105.944; elev=2121 
#minimum and maximum of air and soil temperatures, respectively (C)
Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30 
#wind speed (m/s)
V=1 
#assumptions
tau=0.7; rho=0.6 #atmospheric transmissivity and albedo
Tb0=25 #initial assumption of body temperature (C)

doy= day_of_year("2018-06-01", format= "%Y-%m-%d") 
snoon= solar_noon(lon=lon, doy=doy) #estimate solar noon
dayl= daylength(lat=lat, doy=doy) #estimate daylength
tr= snoon-dayl/2 #time of sunrise
ts= snoon+dayl/2 #time of sunset
```

Although measured solar radiation is preferable if available, we can
estimate hourly solar radiation by discounting incoming solar radiation
as it moves through the atmosphere as follows:

```{r}
#estimate zenith angle (degrees)
psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon) 
#convert to radians
psi_rad= degree_to_radian(psi_deg) 
#estimate radiation
Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho)
#Separate into direct, diffuse, and reflected solar radiation
Sdir= Srad[1,]; Sdif= Srad[2,]; Sref= Srad[3,]
```

We can then calculate hourly air and soil temperatures based on daily
minimum and maximum temperatures. We select the sine-exponential model
for air temperature and the sine model for surface temperature:

```{r}
#air temperature (C)
Ta= sapply(1:24, diurnal_temp_variation_sineexp, T_max=Tmax, T_min=Tmin, t_r=tr, t_s=ts, alpha=2.59, beta= 1.55, gamma=2.2) 
#soil temperature (C)
Ts= sapply(1:24, diurnal_temp_variation_sine, T_max=Tmax_s, T_min=Tmin_s) 
```

We use microclimate models to scale air temperature and wind speed from
weather station height (2m) to lizard height (0.02m). We assume a
surface roughness of $z_0=0.2$, which corresponds to bare sand and
determines the turbulence of airflow. We implement neutral air
temperature and wind speed profiles as well as an air temperature
profile forced by wind speed.

```{r}
#scale air temperature to lizard height
#z is reference height, z is organism height, z0 is surface roughness
Ta_liz= air_temp_profile_neutral(T_r=Ta, zr=2, z0=0.2, z=0.02, T_s=Ts)
Ta_liz2= air_temp_profile(T_r=Ta, u_r=V, zr=2, z0=0.2, z=0.02, T_s=Ts)

#scale wind speed to lizard height
V_liz= wind_speed_profile_neutral(u_r=V, zr=2, z0=0.2,z=0.02)
```

### Energy balance

We will be solving the following energy balance to estimate $T_e$:
$$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$
where $Q_{net}$ is the net energy exchange with the environment (W),
$Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net
thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to
convection (W), $Q_{cond}$ is energy exchange due to conduction (W),
$Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is
the energy emitted due to evaporative water loss (W). We will estimate
each term on the right side of the equation in turn. Estimating
$Q_{abs}$ requires the surface area exposed to radiation and the solar
absorptivity of the animal surface ($a$ proportion). We model a 10g
*Sceloporus* lizard with solar absorptivity $a=0.9$
[@gates1980biophysical]. We will initially assume $T_b=T_a$ to
illustrate the calculations before solving for $T_b$ given the
environmental conditions.

```{r}
mass=10 #mass (g)
a=0.9 #solar absorptivity (%)

#estimate surface area (m^2) from mass (g)
A= sa_from_mass(mass, "lizard")
#estimate projected (silhouette) area as a portion of surface area
psa= sapply(psi_deg, prop_silhouette_area, taxa= "lizard", posture= "prostrate")
#change negative values to zero
psa[psa<0]=0
```

We calculate the hourly solar radiation absorbed (W) as follows:

```{r}
Qabs= rep(NA, 24)
for(hr in 1:24) Qabs[hr]= Qradiation_absorbed(a=a, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], S_dir=Sdir[hr], S_dif=Sdif[hr], a_s=rho)
```

We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. We
assume the surface emissivity of lizards, $epsilon_s=0.965$
[@barlett1967].

```{r}
epsilon_s=0.965

Qemit= rep(NA, 24)
for(hr in 1:24) Qemit[hr]=Qemitted_thermal_radiation(epsilon=epsilon_s, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], T_b=Ta_liz[hr]+273.15, T_g=Ts[hr]+273.15, T_a=Ta_liz[hr]+273.15, enclosed=FALSE)
```

We next estimate convection $Q_{conv}$ (W) and conduction $Q_{cond}$
(W). We will estimate the lizard's heat transfer coefficient, $H_L$
($Wm^{-2}K^{-1}$) using an empirical relationship for lizards
(`heat_transfer_coefficient()`), but also illustrate a function
estimating $H_L$ using a spherical approximation
(`heat_transfer_coefficient_approximation()`) and a simplified
approximation (`heat_transfer_coefficient_simple()`).

```{r}
#Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
ap= airpressure_from_elev(elev)*1000 #estimate Barometric pressure (pascal)
 
DRYAIRout= DRYAIR(db=Ta, bp=ap, alt=elev)
K= DRYAIRout$thcond #thermal conductivity (Wm^-2K^-1) 
nu= DRYAIRout$viskin #kinematic viscosity (m2 s-1)

svl=0.006 #approximate snout vent length (m) for Sceloporus

#We will use the average of K and nu across the day for simplicity and since there is not a substantial diurnal variation.
K= mean(K)
nu=mean(nu)

#Estimate the heat transfer coefficient using an empirical relationship for lizards
H_L=heat_transfer_coefficient(V=V_liz,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6) , taxa="lizard_surface")

#Also illustrate estimations using a spherical approximation and a simplified version of the approximation.
H_L2= heat_transfer_coefficient_approximation(V=V_liz,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6), taxa="lizard")
H_L3= heat_transfer_coefficient_simple(V=V_liz,D=svl)
```

We estimate convective heat exchange between the animal and surrounding
air using the following relationship:
$$Q_{conv}= ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$ where an
enhancement factor, $ef$, multiplier can be incorporated to account for
increases in heat exchange resulting from air turbulence in field
conditions. We implement the function in R assuming that 2/3 of the
lizard's surface area is exchanging heat through convection. We note we
currently estimate no convection since we are assuming $Tb=Ta$ for
illustration purposes:

```{r}
Qconv= rep(NA, 24)
for(hr in 1:24) Qconv[hr]= Qconvection(T_a= Ta_liz[hr]+273.15, T_b= Ta_liz[hr]+273.15, H=H_L, A=A, proportion=0.67, ef=1.3)
```

We estimate conductive heat flow (W) from the lizard to the surface
assuming conductance through the animal tissue is the rate limiting step
as follows: $$Q_{cond}= A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
We implement the estimate in R assuming that 1/3 of the lizard surface
is in contact with the ground and a skin thickness $d$ of $10^{-6}$:

```{r}
Qcond= rep(NA, 24)
for(hr in 1:24) Qcond[hr]=Qconduction_animal(T_g= Ts[hr]+273.15, T_b=Ta_liz[hr]+273.15, d=10^-6, K=K, A=A, proportion=0.33)
```

We assume, as in generally done for lizards, that heat exchange
associated with metabolism and evaporation is negligible:

```{r}
Qmet= 0; Qevap= 0
```

The full heat budget can be calculated as follows
[@gates1980biophysical]:

```{r}
Qnet= Qnet_Gates(Qabs=Qabs, Qemit=Qemit, Qconv=Qconv, Qcond=Qcond, Qmet=Qmet, Qevap=Qevap)
```

We constructed the energy balance assuming $T_b=T_a$ for illustration.
Thus, the energy balance is dominated by heat gain from contact with the
warmer surface. We now use a function based on the energy balance above
to estimate body temperature given the environmental conditions:

```{r}
Te= rep(NA, 24)
for(hr in 1:24) Te[hr]= Tb_Gates(A=A, D=svl, psa_dir=psa[hr], psa_ref=1-psa[hr], psa_air=0.67, psa_g=0.33, T_g=Ts[hr]+273.15, T_a=Ta_liz[hr]+273.15, Qabs=Qabs[hr], epsilon=epsilon_s, H_L=H_L, ef=1.3, K=K)
```

We also implement a similar but simplified energy balance
[@campbell2000introduction]. The energy balance omits conduction with
the ground:

```{r}
Te2= rep(NA, 24)

for(hr in 1:24) Te2[hr]= Tb_CampbellNorman(T_a=Ta_liz[hr]+273.15, T_g=Ts[hr]+273.15, S=Qabs[hr]/A, alpha_L=0.96, epsilon=epsilon_s, c_p=29.3, D=svl, V=V_liz)
#S is solar radiation flux (W m^-2), so we divide by surface area, A
```

We additionally estimate $T_e$ using a specialized function for lizards
[@buckley2008link]:

```{r}
Te3= rep(NA, 24)
for(hr in 1:24) Te3[hr]= Tb_lizard(T_a=Ta_liz[hr], T_g=Ts[hr], u=V_liz, svl=svl*1000, m=mass, psi=psi_deg[hr], rho_S=rho, elev=elev, doy=doy, sun=TRUE, surface=TRUE, alpha_S=a, alpha_L=0.965, epsilon_s=epsilon_s, F_d=0.8, F_r=0.5, F_a=0.5, F_g=0.5)
```

### Plot

We next plot the diurnal trends in environmental data and predicted body
temperatures.

```{r, echo=FALSE}
par(mar=c(5, 4, 4, 8) + 0.1)
plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab="Temperature (°C)", col="#5DC863FF", ylim=c(10,60)) #Gates
points(1:24, Te2-273.15, type="l", col="#5DC863FF", lty="dotted") #Campbell and Norman
points(1:24, Te3, type="l", col="#5DC863FF", lty="dashed") #Lizard, Buckley 2008
points(1:24, Ta, type="l", col="#440154FF")
points(1:24, Ta_liz, type="l", col="#3B528BFF")
points(1:24, Ts, type="l", col="#21908CFF")

#add additional axis with radiation
par(new = T)
plot(1:24, Sdir, pch=16, axes=F, xlab=NA, ylab=NA, type="l", col="#FDE725FF")
axis(side = 4)
mtext(side = 4, line = 3, 'Radiation (W/m^2)')
legend("topleft",
       legend=c("Ta 2m", "Ta 0.02m", "Ts", "Te", "Radiation"),
       lty=1, pch=NA, col=c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF"))

```

*Figure 1*. Predicted lizard body temperatures (Te) as a function of air
temperatures (Ta) at 2m and lizard height (0.02m), surface (Ts)
temperatures (°C), and solar radiation ($W/m^2$). We depict body
temperatures using two general energy budgets [solid: Tb_Gates();
dotted: Tb_CampbellNorman()] and a lizard specific biophysical model
[dashed: Tb_lizard()].

The microclimate models indicate that air temperatures at lizard height
are similar to surface temperatures (Figure 1). The biophysical models
indicate that solar radiation will elevate lizard body temperatures far
above air temperatures and that the lizard will face thermal stress if
it is unable to seek shade (Figure 1). The three biophysical models
predict different body temperatures during peak period of solar
radiation because they model interactions with the ground differently
and users are encouraged to review the details of each biophysical model
before selection. Differences in estimated body temperatures are
accentuated by the high level of solar radiation.

## Conclusions

TrenchR is intended to promote understanding of how organisms interact
with their environment and the consequences for physiology, energetics,
behavior, and demography. Our example implementation highlights the
importance of considering organismal body temperatures, rather than air
temperatures, when examining thermal stress and other responses to
environmental variability and change. TrenchR currently focuses on heat
balances but may be expanded to include water balances. Simple functions
can be combined as needed to produce comprehensive and transparent
models for biophysical ecology and evolution. The resultant models are
likely to be sufficiently detailed for many applications, but users are
referred to NicheMapR for more detailed biophysical models
[@kearney2017nichemapr; @kearney2020nichemapr]. Making classic
biophysical ecology techniques more accessible will allow researchers to
take advantage of rapidly accumulating data on environmental conditions
and organismal traits to understand and predict ecological and
evolutionary responses to variable and changing environments.

## Acknowledgements

We thank Brian Helmuth and Eric Riddell for providing and checking
functions. We thank Ray Huey, Joel Kingsolver, Luke Miller, Warren
Porter, and Art Woods for assistance and input. We particularly thank
Michael Kearney for extensive input on TrenchR and its alignment with
NicheMapR and for contributing two educational modules.

[Check with Eric]

## Authors' contributions

LB conceived and designed the package and lead all aspect of the
packages creation and dissemination; BBO, IC, AJ, OL, AM, YS wrote and
edited functions and associated package resources; AJ configured the
package; AM, BBO, YS, and LB converted and edited the educational
modules. All authors contributed critically to the drafts and gave final
approval for publication.

## Data availability

All code is available at <https://github.com/trenchproject/TrenchR>.

## References
