---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan Briones Ortiz, Aji John, Ofir Levy, Yutaro Sakairi,..."
date: "February 4, 2020"
output:
  word_document: default
  html_document: default
  pdf_document: default
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

Much understanding of climate change responses relies on the assumption that organismal body temperatures are equal to temporally averaged air temperatures high above the ground. However, most organisms live near the ground and acute exposure to solar radiation can substantially elevate their body temperatures. We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses (https://github.com/trenchproject/TrenchR). The package includes microclimate models to vertically scale weather station data to organismal heights. The biophysical modelling tools include both general models for heat flows and specific models to predict body temperatures for a variety of ectothermic taxa. Additional functions model and temporally partition air and soil temperatures and solar radiation. The package aims to introduce and enable microclimate and biophysical modelling to improve ecological and evolutionary forecasting and includes tutorials and well as a series of educational modules.

## Background
Many responses of species and ecosystems to climate change are inconsistent with predictive most current models [@maguire2015modeling]. Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic]. Many analyses assume that organismal body temperatures are equal to air temperate at weather station height (usually 2m). However, air temperatures near the ground where most organisms reside can be considerably warmer and absorption of solar radiation can raise body temperatures well above air temperatures, resulting in potential discrepancies of tens of degrees C [Kearney et al., 2009; Sinclair et al., 2016; Sunday et al., 2014].  Acute thermal stress events are often more relevant to the physiology and energy balance of organisms than daily, monthly, or annually aggregated environmental conditions. Yet, daily and seasonal environmental variation is often neglected [@dillon2016introduction; @dillon2016life]. We aim to help remedy these shortcomings by introducing the TrenchR package as a component of the TrEnCh project (https://trenchproject.github.io), which builds computational and visualization tools to Translate Environmental Change into organismal responses.

The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books [@campbell2000introduction; @gates1980biophysical] and articles [@porter1969thermodynamic; @porter1983biophysical], but adoption of the tools has not kept pace with research on climate change responses. Adoptions of biophysical approaches are expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses [@buckley2018leveraging; @kearney2017nichemapr].  Increasing dissemination of computational tools is facilitating adoption. The TrenchR package is intended to complement the NicheMapR package [@kearney2017nichemapr], which includes sophisticated microclimate and biophysical models. Many of the approaches are similar between the packages, but TrenchR focuses on simple functions to aid the understanding and transparency of biophysical approaches.  The TrenchR functions allow for modular use and easy adaptation to a variety of organismal and research questions.  Some energy budget calculation components of biophysical models are available in the ThermImage package (Tattersall, 2017). Although many of the TrenchR functions are general, some components of TrenchR are best suited for ectothermic animals. We have minimized inclusion of plant specific biophysical models since they are the focus of several R packages [tealeaves: (Muir, 2019); plantecophys [@duursma2015plantecophys]].

## Methods and features
The TrenchR package aims to promote transparency and reproducibility and we welcome contributions and corrections from users. Our package is built using the devtools methodology (https://github.com/hadley/devtools) in Github. Issues and feature requests can be contributed in Github (https://github.com/trenchproject/TrenchR/). The package is maintained and checked using Travis CI as a tool for continuous integration. 

Functions were adapted from biophysical ecology textbooks [@campbell2000introduction; @gates1980biophysical] and research articles. The source reference is included in each function’s header. We have incorporated default parameters (e.g., organism emissivity or heat conduction rates) when general values are available. Many functions additionally include comments that describe potential parameterizations. The biophysical ecology texts [@campbell2000introduction; @gates1980biophysical] include many tables with parameter values and we provide several tables in TrenchR [Table 8.2XXXX].

### Components
TrenchR functions are organized into the following categories:
**Utility functions** allow estimating environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for development. 

**Microclimate functions** allow estimating the environmental conditions experienced by organisms. Temperature and wind profile functions scale environmental conditions from sensors to organismal height. Radiation functions allow estimating incoming solar radiation and partitioning the variation diurnally and across components (i.e., direct, diffuse, reflected).

**General biophysical functions** provide generalized models of heat exchanges between organisms and their environment so that users can build custom biophysical models. The functions allow implementing an energy balance of the following components of heat exchange between organisms and their environment: *radiative heat exchanges* of solar and thermal radiation; *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; *conductive heat exchanges* between organisms and the ground due to physical contact; and *evaporative heat exchanges* associated with organisms’ evaporative water loss. Additional functions aggregate these forms of heat exchange into energy balances and to use the energy balances to predict body temperatures. 

**Specific biophysical functions** provide biophysical models that have been built for specific organisms based on their physical properties, behavior, environment, and life history. We currently provide published biophysical models for lizards, salamanders, butterflies, grasshoppers, limpets, mussels, and snails. Most models predict operative environmental temperatures, which are the equilibrium body temperatures of organisms with specified physical properties in a specific microenvironment and assume no heat exchange via metabolism or evaporation [@bakken1992measurement].

### Vignettes and introductory modules
We introduce the function in several vignettes.  A good place to start is the *Allometry and conversions* tutorial, which provides tools for preparing data such as estimating additional dimensions of organisms from measured dimensions. The *Estimating microclimates* tutorial provides resources for estimating the environmental conditions experienced by organisms.  This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and windspeed profiles, and soil temperatures and profiles. Finally, the core biophysical modelling functions are described in a tutorial on *Using energy balances to estimate body temperatures*. Components of an energy budget can be estimated using individual functions and then body temperatures can be solved for using either a generic energy balance or taxa specific biophysical models.

For additional background on microclimate and biophysical modelling, we have updated a series of educational modules entitled “Physical processes in ecosystems”.  We intend the modules to provide a more contained and accessible introduction to microclimate and biophysical modelling than that included in classic textbooks [@campbell2000introduction; @gates1980biophysical]. The modules provide less detailed coverage of many of the topics contained in the books. The series of 14 modules starts with an overview of the calculus and physics principles underlying the modelling. Thermodynamics and energy budget modelling are then introduced with detailed examples. Modules address the climate space concept, operative temperatures, biophysical models for leaves and sheep, and heat flow in soils. We have revised the modules (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R code and utilization of TrenchR functions. We expanded the original series of tutorials to include introductions by Michael Kearney to the Microclim environmental data [@kearney2014microclim] and integrating the Microclim data with biophysical modelling to examine broad scale climatic limits. The modules are provided in html form and available for download as a pdf at the bookdown server (https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/). The R markdown files are available in Github at https://github.com/trenchproject/TrenchRmodules.  

## Examples
We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2018, in Santa Fe, New Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also incorporated in the *Using energy balances to estimate body temperatures* tutorial. We first estimate the energy budget with component functions and we then use an integrated biophysical model to estimate operative environmental temperatures. We start by generating environmental inputs.

We will use the energy budget to estimate the lizard's body temperature, $T_b$, which can depart dramatically from the air temperatures due to energy exchange with the environment.  Heat is gained from absorbing solar and thermal radiation and from metabolic reactions. Heat is lost through the organism's emission of radiation and the evaporation of water. The organism exchanges heat with the surrounding air or water via convection and with the ground via conduction. The balance of these heat exchanges can be estimated and often referred to as operative environmental temperature, $T_e$ [@bakken1992measurement].

Let's assume the lizard is in an unshaded location where the daily air temperature varies from a minimum of 10°C to a maximum of 25°C, the soil surface temperature varies from a minimum of 15°C to a maximum of 30°C, and the windspeed is 0.5 m/s. We assume that atmospheric transmissivity $\tau=0.7$ and albedo $\rho=0.6$

### Environmental data

Let us first prepare the environmental data for analysis. We will need to estimate hourly air and soil temperatures and radiation. We start by estimating the day of year and the timing of sunrise and sunset:
```{r}
library(TrenchR)

#set up input data as variables
lat=35.69; lon= -105.944; elev=2121 #latitude and longitude (degrees) and elevation (m)
Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30 #minimum and maximum of air and soil temperatures, respectively (C)
V=0.5 #wind speed (m/s)
#assumptions
tau=0.7; rho=0.6 #atmospheric transmissivity and albedo
Tb0=25 #initial assumption of body temperature (C)

doy= day_of_year("2018-06-01", format= "%Y-%m-%d") 
snoon= solar_noon(lon=lon, doy=doy) #estimate solar noon
dayl= daylength(lat=lat, doy=doy) #estimate daylength
tr= snoon-dayl/2 #time of sunrise
ts= snoon+dayl/2 #time of sunset
```

We can estimate hourly solar radiation as follows:
```{r}
psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon) #estimate zenith angle (degrees)
psi_rad= degree_to_radian(psi_deg) #convert to radians
Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho) #estimate radiation
#Separate into direct, diffuse, and reflected solar radiation
Sdir= Srad[1,]
Sdif= Srad[2,]
Sref= Srad[3,]
```

We can then calculate hourly air and soil temperatures. We use the sine-exponential model for air temperature and the sine model for surface temperature:
```{r}
Ta= sapply(1:24, diurnal_temp_variation_sineexp, T_max=Tmax, T_min=Tmin, t_r=tr, t_s=ts, alpha=2.59, beta= 1.55, gamma=2.2)
Ts= sapply(1:24, diurnal_temp_variation_sine, T_max=Tmax_s, T_min=Tmin_s)
```

### Energy balance
We will be solving the following energy balance to estimate $T_e$:
$$Q_{net} = Q_{abs} - Q_{emit} + Q_{conv} + Q_{cond} + Q_{met} - Q_{evap},$$
We will estimate each term on the right side of the equation in turn. Estimating $Q_{abs}$requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion). We model a 10g Sceloporus lizard with solar absorptivity $a=0.9$ [@gates1980biophysical]. We will initially assume $T_b=T_a$ to illustrate calculations.

```{r}
mass=10
a=0.9 #solar absorptivity

#estimate surface area (m^2) from mass (g)
A= sa_from_mass(mass, "lizard")
#estimate projected (silhouette) area as a portion of surface area
psa= sapply(psi_deg, prop_silhouette_area, taxa= "lizard", posture= "prostrate")
#change negative values to zero
psa[psa<0]=0
```

We calculate the hourly radiation absorbed as follows:
```{r}
Qabs= rep(NA, 24)
for(hr in 1:24) Qabs[hr]= Qradiation_absorbed(a=a, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], S_dir=Sdir[hr], S_dif=Sdif[hr], a_s=rho)
```

We estimate thermal radiation $Q_{emit}$ for the lizard outdoors. We assume the surface emissivity of lizards, $epsilon_s=0.965$ [Bartlett & Gates 1967].

```{r}
epsilon_s=0.965

Qemit= rep(NA, 24)
for(hr in 1:24) Qemit[hr]=Qemitted_thermal_radiation(epsilon=epsilon_s, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], T_b=Ta[hr]+273.15, T_g=Ts[hr]+273.15, T_a=Ta[hr]+273.15, enclosed=FALSE)
```

We next estimate convection $Q_{conv}$ and conduction $Q_{cond}$. We will estimate the lizard's heat transfer coefficient, $H_L$ using an empirical relationship for lizards (`heat_transfer_coefficient()`), but also illustrate a function estimating $H_L$ using a spherical approximation (`heat_transfer_coefficient_approximation()`) and a simplified approximation (`heat_transfer_coefficient_simple()`).  

```{r}
#Use DRYAIR in NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
#library(NicheMapR)

#estimate Barometric pressure (pascal)
ap= airpressure_from_elev(elev)*1000
  
DRYAIRout= DRYAIR(db=Ta, bp=ap, alt=elev)
K= DRYAIRout$thcond
nu= DRYAIRout$viskin #Kinematic viscosity (m2 s-1)

svl=0.006 #SVL for sceloporus

#We will use the average of K and nu across the day for simplicity and since there's not a substantial variation
K= mean(K)
nu=mean(nu)

#Estimate using an empirical relationship
H_L=heat_transfer_coefficient(V=V,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6) , taxa="lizard_surface")

#Also illustrate estimations using a spherical approximation and a simplified version of the approximation.
heat_transfer_coefficient_approximation(V=V,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6), taxa="lizard")
heat_transfer_coefficient_simple(V=0.5,D=0.05)

```

We estimate convective heat exchange between the animal and surrounding air using ithe following relationship:
$$Q_{conv}= ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$
where an enhancement factor, $ef$, multiplier can be incorporated to account for increases in heat exchange resulting from air turbulance in field conditions

The function is available in R assuming that 2/3 of the lizard's surface area is exchanging heat through convection. We note we currently estimate no convection since we are assuming $Tb=Ta$ for illustration purposes:
```{r}
Qconv= rep(NA, 24)
for(hr in 1:24) Qconv[hr]= Qconvection(T_a= Ta[hr]+273.15,T_b= Ta[hr]+273.15,H=H_L,A=A, proportion=0.67, ef=1.3)
```

We estimate conductance between the lizard and surface assuming conductance through the animal tissue is the rate limiting step as follows:
$$Q_{cond}= A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d, $$
We conduct the estimate in R assuming that 1/3 of the lizard surface is in contact with the ground and a skin thinkness of $10^{-6}$:
```{r}
Qcond= rep(NA, 24)
for(hr in 1:24) Qcond[hr]=Qconduction_animal(T_g= Ts[hr]+273.15,T_b=Ta[hr]+273.15,d=10^-6,K=K,A=A, proportion=0.33)
```

We assume, as in generally done for lizards, that heat exchange associated with metabolism and evaporation is negligible:
```{r}
Qmet= 0
Qevap= 0
```

The full heat budget can be calculated as follows: 
```{r}
Qnet_Gates(Qabs=Qabs, Qemit=Qemit, Qconv=Qconv, Qcond=Qcond, Qmet=Qmet, Qevap=Qevap)
```
We constructed the energy balance assuming $T_b=T_a$ for illustration. We now use a function based on the energy balance above to estimate body temperature given the environmental conditions: 

```{r}
Te= rep(NA, 24)
for(hr in 1:24) Te[hr]= Tb_Gates(A=A, D=svl, psa_dir=psa[hr], psa_ref=1-psa[hr], psa_air=0.67, psa_g=0.33, T_g=Ts[hr]+273.15, T_a=Ta[hr]+273.15, Qabs=Qabs[hr], epsilon=epsilon_s, H_L=H_L, ef=1.3, K=K)
```

We also implement a similar but simplified energy balance. The energy balance omits conduction with the ground and thus underestimates temperatures, but we implement it for comparison purposes:
```{r}
Te2= rep(NA, 24)
for(hr in 1:24) Te2[hr]= Tb_CampbellNorman(T_a=Ta[hr]+273.15, S=Qabs[hr]/svl, epsilon=epsilon_s, c_p=29.3, D=svl, V=V)
#CHECK
```

We additionally estimate $T_e$ using the specialized function for lizards:
```{r}
Te3= rep(NA, 24)
for(hr in 1:24) Te3[hr]= Tb_lizard(T_a=Ta[hr], T_g=Ts[hr], u=V, svl=svl*1000, m=mass, psi=psi_deg[hr], rho_S=rho, elev=elev, doy=doy, sun=TRUE, surface=TRUE, alpha_S=a, alpha_L=0.965, epsilon_s=epsilon_s, F_d=0.8, F_r=0.5, F_a=0.5, F_g=0.5)

Te4= rep(NA, 24)
for(hr in 1:24) Te4[hr]= Tb_Fei(T_a=Ta[hr]+273.15, T_g=Ts[hr]+273.15, H=Qabs[hr], lw=Qabs[hr], shade=0, m=mass, Acondfact =0.1, Agradfact =0.3)
#CHECK

```

### Plot
We next plot the diurnal trends in environmental data and the operative data.

```{r}
par(mar=c(5, 4, 4, 8) + 0.1)
plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab="Temperature (°C)", col="blue") #Gates
points(1:24, Te2-273.15, type="l", col="blue", lty="dotted") #Campbell
points(1:24, Te3, type="l", col="blue", lty="dashed") #Buckley 2008
#points(1:24, Te4, type="l", col="red", lty="solid") #Fei et al.
points(1:24, Ta, type="l", col="orange")
points(1:24, Ts, type="l", col="purple")

#add additional axis with radiation
par(new = T)
plot(1:24, Sdir, pch=16, axes=F, xlab=NA, ylab=NA, type="l", col="green")
axis(side = 4)
mtext(side = 4, line = 3, 'Radiation')
legend("topleft",
       legend=c("Ta", "Ts", "Te", "Radiation"),
       lty=1, pch=NA, col=c("orange", "purple", "blue", "green"))

```

## Summary
New environmental data and tools make biophysical modeling more accessible. Expand use since we realize.

## References
